Why do languages tolerate heterography? An experimental investigation into the emergence of informative orthography

### 2024-04-09 manuscript/manuscript.tex
Second revision

% temp for revision %%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{xcolor}
\newcommand\newmaterial[1]{\textcolor{blue}{#1}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\authorsnames{Anonymous}

\authorsaffiliations{{Anonymous}}

@@

Despite the benefits that heterography may provide in reading, it comes with two main costs. Firstly, by definition, heterography implies that a single sound can be spelled multiple ways. In English, the heterographic spellings <meat> and <meet> imply that /i/ can be spelled <ea> or <ee>. Readers are therefore required to learn alternate spellings for a single sound, \newmaterial{resulting in longer learning periods and more difficult decoding \parencite{Reis:2020, Seymour:2003, Spencer:2003, Taylor:2011, Zhao:2018}}. Secondly, the arbitrary mapping between heterographic forms and meaning must also be learned. From the point of view of a modern English speaker, there is no intrinsic reason why \textit{meat} is spelled <ea> and \textit{meet} is spelled <ee>. Nevertheless, these arbitrary spelling distinctions must be learned if they are to be useful, and they presumably place an additional burden on reading and---perhaps even more-so---on writing \parencite{Frith:1979, Shankweiler:1992}.

@@

First, however, it is useful to consider the mechanisms by which selection could occur if it is indeed the case that heterography emerges for functional reasons. \textcite[pp.~325--326]{Berg:2021} outline two models of how a word might enter a state of heterography. The first model, \textit{the differentiation model} (Fig.~\ref{fig01}A), explains heterography through the creation of new orthographic forms. For example, the spelling <lite> for the word \textit{light} is frequently used in food products to mean light-in-calorific-weight; in British English, the spelling <cheque> (\newmaterial{perhaps influenced by French \textit{chèque}}) differentiates the bank draft from other meanings of the word \textit{check}; and the word \textit{byte} was a deliberate respelling of \textit{bite} to avoid accidental mutation into the closely related term \textit{bit} \parencite{Buchholz:1977}. Many monosyllabic words that are homophonous with common function words also tend to adopt alternate spellings, often by adding <e> or by doubling the final consonant: \textit{be}--\textit{bee}, \textit{but}--\textit{butt}, \textit{by}--\textit{bye}--\textit{buy}, \textit{for}--\textit{fore}--\textit{four}, \textit{in}--\textit{inn}, \textit{or}--\textit{oar}--\textit{ore}, \textit{so}--\textit{sew}, \textit{to}--\textit{too}--\textit{two}, \textit{we}--\textit{wee}. Differentiation is also common in surnames---\textit{Clarke}, \textit{Greene}, \textit{Wilde}; \textit{Carr}, \textit{Hogg}, \textit{Mann}---and trade names---\textit{Blu Tack}, \textit{Froot Loops}, \textit{Wite-Out} \parencite[§~6]{Carney:1994}. One important way in which differentiation can occur---especially in a language like English that has historically contained a lot of spelling variation \parencite{Nevalainen:2012, Stenroos:2016}---is by the conditioning of variant spellings on meaning; pairs like \textit{discreet}--\textit{discrete}, \textit{flour}--\textit{flower}, and \textit{plain}--\textit{plane}, which were once variant spellings of the same word, have taken on distinct meanings over time \parencite[§~5.4]{Carney:1994}. \textcite[p.~58]{Berg:2017} have referred to this as the ``functionalization of leftovers'': The spelling variants that survive are those that ``can find distributional or functional niches.''

@@

Our paper reports the results of two experiments---focusing on the differentiation and conservation models respectively---with the goal of demonstrating that functional heterography arises preferentially under a communicative need for disambiguation. All data and code is available from \url{https://osf.io/7auw6/?view_only=f8c2498b4add41839f51bdbefa03392e}. \newmaterial{To increase the transparency of our work, we created a preregistration at \url{https://aspredicted.org/YDV_MXV}. Note, however, that due to the more exploratory nature of this project, our preregistration did not specify strong confirmatory hypotheses or precise statistical models, focusing instead on the research question, experimental conditions, general predictions, primary measurement constructs, sample size, and exclusion criteria.}

@@

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/fig07.pdf}}
	\vspace*{2pt}
	\caption{Results of Experiment~1. \textbf{A} Typological distribution by generation and condition over the four typological categories: holistic (H; purple), expressive (E; blue), redundant (R, green), and degenerate (D; yellow). \textbf{B} Proportion of systems classified as informative (holistic or expressive) by generation. The dots show the observed proportions and the curves show logistic regression models fit to the data. \textbf{C} Communicative cost by generation along with regression models fit to the data. \textbf{D} Transmission error by generation along with regression models fit to the data. The panels on the right show the posterior estimates of the intercept ($\alpha$) and slope ($\beta$) parameters by condition as well as the posterior differences between conditions. The green, blue, and red bars indicate respectively the 95\%, 90\%, and 85\% HDIs (credible intervals).}
	\label{fig07}
	\end{figure}

@@

\newmaterial{To analyze how informativeness changes over time and how this compares between the two conditions, we reduced the typological classifications into two broader categories: informative systems (i.e., holistic or expressive) and uninformative systems (i.e., redundant or degenerate). We then fit a Bayesian mixed-effects logistic regression model that predicts whether or not a system is informative as a function of generation with by-chain random slopes and intercepts,\footnote{Model: informative $\sim$ log(generation) + (1 + log(generation) | chain). Bernoulli likelihood with logit link function, weakly informative default priors, six chains of 12000 samples. All statistical models were fit using Bambi 0.13 \parencite{Capretto:2022}. Parameter estimates and diagnostics are provided in Appendix~B of the supplementary material.} which is the standard model structure used to analyze iterated learning experiments \parencite{Winter:2016}. Our Bayesian approach produces posterior estimates of two key parameters: $\alpha$, representing the intercept of the regression model, and $\beta$ representing its slope. To determine if there is a statistical difference between conditions, we compute the difference in slopes, $\Delta(\beta) = \beta_\mathrm{comm} - \beta_\mathrm{trans}$, and check if this posterior difference satisfactorily rejects zero. Here we follow the convention that the 95\% highest density interval (HDI; the narrowest interval that contains 95\% of the posterior probability mass) should not include zero. We emphasize, however, that the posterior is a complete description of the evidence (given the data and model assumptions) and does not strictly need to be reduced to a binary yes/no decision. The results are shown in Fig.~\ref{fig07}B. In both conditions, there is a decrease in informativeness over time (the slopes are negative), but in the Transmission + Communication condition, the slope is shallower, suggesting that informativeness decreases more slowly in the presence of communicative pressure. The difference in slopes, $\Delta(\beta) = 4.53$ (95\%~HDI: 1.48, 7.88), clearly rejects zero, pointing to a meaningful difference between conditions.}

\newmaterial{One issue with the above approach is that collapsing the systems into binary categories (informative vs.\@ uninformative) results in a loss of information about \textit{how informative} the systems are. In addition, there was evidence to suggest that the model was a suboptimal description of the data, since there was also a difference in the $\alpha$ estimates (the intercepts), which should theoretically be the same (i.e., there should be no difference between conditions at Generation~0). We address these limitations with a second measure of informativeness, communicative cost \parencite{KempRegier:2012, Kemp:2018, Regier:2015}, which has previously been used in similar experimental studies \parencite{Carr:2020, Carstensen:2015, Smith:2020} and was proposed in our preregistration as the primary measure of informativeness.} Communicative cost is an information-theoretic measure that expresses how much information will be lost, on average, when a speaker/writer attempts to convey a meaning to a listener/reader using some shared signaling system. If the system contains no ambiguities (all meanings are expressed by unique signals), communicative cost will be zero~bits---that is, zero information will be lost during each attempt to communicate using that system. Communicative cost will take some larger value if the system contains ambiguity. It is given by $\sum_{m \in U} \mathrm{Pr}(m) ( -\log \mathrm{Pr}(m|s_m) )$, where $U$ is the universe of meanings that may be expressed, $\mathrm{Pr}(m)$ is the probability that a particular meaning would need to be expressed, and $\mathrm{Pr}(m|s_m)$ is the probability that a reader would infer meaning $m$ given that a writer produced signal $s$ for meaning $m$. In our case, $U$ is the set of nine alien objects, $\mathrm{Pr}(m)$ is set to $1/|U|$ (all objects need to be talked about with equal probability), and $\mathrm{Pr}(m|s_m)$ is given by $1/|M_s|$, where $M_s$ is the set of meanings labeled $s$ according to the system. \newmaterial{The results are shown in Fig.~\ref{fig07}C. We used the same mixed-effects linear regression model described above (except that the likelihood is now Gaussian). In line with the previous analysis, cost increases with generation in both conditions but increases more slowly under communicative pressure. However, the support for a difference between conditions was weaker under this more nuanced measure, $\Delta(\beta) = -0.17$ (95\%~HDI: $-0.36$, $0.02$); although we were not able to reject zero at the 95\% level, we were able to reject it at the 92\% level (92\%~HDI: $-0.35$, $-0.01$).}

\newmaterial{Aside from the informativeness of the orthographic systems, we also predicted in our preregistration that the systems would become easier to learn over time in both conditions, albeit for slightly different reasons. In Transmission-only, the orthographic system is expected to become increasingly learnable as it degenerates into a single, transparent suffix form. In Transmission~+ Communication, the system is expected to become more learnable as the unsystematic, holistic systems transform into other easier to learn systems (notably expressive systems, although---as noted already---expressive systems rarely emerged). Following prior work, we operationalized learnability as transmission error---the amount of error that the participant at Generation~$i$ made in reproducing the orthographic system that existed at Generation~$i-1$; transmission error is defined as the mean Levenshtein edit distance between the corresponding orthographic forms in consecutive generations \parencite[see e.g.,][]{Kirby:2008}. These results are plotted in Fig.~\ref{fig07}D. In both conditions, the estimates of the $\beta$ parameters are negative and clearly reject zero, suggesting that the systems do indeed become increasingly learnable over time. There was some weak evidence for a difference in the intercepts, with transmission error remaining higher in the communicative condition. This is to be expected because the goal in the Transmission-only condition is to reproduce the forms taught in training, whereas the goal in the communicative condition is to devise a system that permits accurate communication, which necessitates more change to the system that was learned.}

@@

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/fig11.pdf}}
	\vspace*{2pt}
	\caption{Results of Experiment~2. \textbf{A} Typological distribution by generation and condition over the four typological categories: holistic (H; purple), expressive (E; blue), redundant (R, green), and degenerate (D; yellow). \textbf{B} Proportion of systems classified as informative (holistic or expressive) by generation. The dots show the observed proportions and the curves show logistic regression models fit to the data. \textbf{C} Communicative cost by generation along with regression models fit to the data. \textbf{D} Transmission error by generation along with regression models fit to the data. The panels on the right show the posterior estimates of the intercept ($\alpha$) and slope ($\beta$ and $\gamma$) parameters by condition as well as the posterior differences between conditions. $\beta$ represents the effect of epoch and $\gamma$ represents the effect of generation number within epoch. The green, blue, and red bars indicate respectively the 95\%, 90\%, and 85\% HDIs (credible intervals).}
	\label{fig11}
	\end{figure}

@@

Fig.~\ref{fig11}A plots the typological distributions by generation and condition. Initially, all systems are expressive, but the dominance of this category is gradually eroded over time, particularly during the second and third epochs once the spoken forms had become homophonous. Notably, however, the loss of expressive systems appears to be slower in Transmission~+ Communication, and redundant systems were also less popular under communicative pressure. \newmaterial{As in Experiment~1, we further collapsed the typological categories into two broader categories (informative vs.\@ uninformative) to analyze the trends over time. The results, shown in Fig.~\ref{fig11}B, show that the probability of a system being informative drops over time in both conditions (primarily as a function of epoch), but does so more slowly in the communicative condition. Although there was some weak evidence of a difference in epoch slopes ($\Delta(\beta) = 2.65$; 95\%~HDI: $-0.58$, $5.95$), we could not conclusively reject zero under this measure.}

\newmaterial{The results in terms of the preregistered measure of informativeness, communicative cost, are presented in Fig.~\ref{fig11}C. Here we find a nonzero effect of both epoch and generation on cost, as well as a difference between conditions in terms of epoch: $\Delta(\beta) = -0.2$ (95\%~HDI: $-0.4$, $-0.003$). Like the previous analysis, the $\gamma$ slopes were in close alignment. This suggests that the effect of generational turnover is very similar between conditions and that the difference between conditions is mostly driven by the increases in homophony induced in each epoch. The overall result is that, in Transmission~+ Communication, the increase in cost is linear across the nine generations, whereas in the Transmission-only condition, the increase in cost follows something more akin to a step function, with sudden increases in cost in response to each additional bout of homophony. In other words, in the Transmission-only condition, the orthographic systems respond rapidly to the changing spoken forms, while in the Transmission~+ Communication condition, the orthographic systems are more resistant to the homophony.}

@@

Although the emergent orthographies tended to be slightly more informative under communicative pressure, systematic differentiation was, on the whole, rare, unstable, and fleeting, be it through implicit generalization of the supplied variants or explicit innovation of new variants. This is \textit{not} because participants were unable to learn variant spellings; indeed, in many cases variant spellings were retained but ineffectually conditioned on shape rather than color. Nor was it because participants were unable to learn a system of color marking that is not expressed in the spoken language; we know from Experiment~2 that participants can learn and reproduce such systems. Instead, it seems that, in Experiment~1, differentiation could not get off the ground. We see two reasons for this. First, participants seemed disinclined to directly encode meaning in how they chose to spell, preferring instead to ``write by ear'' \parencite{Frith:1979}. When asked to type in a word for a pink pentagon called /buvɪkəʊ/, participants were inclined to type a sequence of graphemes that reflected the sound they heard, without encoding meaning. \newmaterial{This behavior might be connected to the concept of ``functional fixedness'' \parencite{German:2000}, which states that learners find it difficult to adduce a new function (e.g., writing meaning) when they are accustomed to another function (e.g., writing sound), which highlights a potentially important role for generational turnover in the development of writing systems, since new learners will be more receptive to new functions.} Second, even when participants \textit{did} appreciate the need to differentiate the written forms to be successful, they often appeared reluctant or unable to do so, perhaps because they viewed the spelling as immutable or because the problem of aligning with a partner---even in a~synchronous setting---was too difficult to overcome \newmaterial{without the ability to coordinate over an extended period of time}. It is notable, for example, that in the communication games, many attempts to differentiate using English color letters were not reciprocated.

@@

\newmaterial{During the review process, a concern was raised that our participants might not have fully understood the communicative nature of the task, thus explaining why we did not observe the emergence of systematic differentiation in spelling. Our position, as outlined above, was that the lack of differentiation was due to the very strong homophony pressure. To test whether the lack of informativeness might be attributed to the homophony and to check that the design of our experiment and implementation of the communicative pressure was sufficient to promote more informative systems, we ran an additional experiment. This experiment was identical to the communicative condition of Experiment~1, except that we removed the spoken forms (thereby replicating previous iterated learning experiments, which are generally only orthographic in nature; e.g., Beckner et al., 2017; Kirby et al., 2008, 2015) and altered the orthographic forms so that no homophony was implied in the spellings. This experiment is described in Appendix~C of the supplementary material, but in short, we observed much greater levels of innovation and informativeness, and a commensurate increase in communicative success (a point we return to shortly). Importantly, this suggests that the participants in Experiment~1 did indeed understand the communicative imperative, but nevertheless preferred their spellings to encode sound rather than meaning. Or, rather, the cultural evolutionary process ultimately tended to favor simplicity over informativeness in this particular domain.}

@@

@@

\newmaterial{The results of both experiments were relatively weak statistically, with the differences in terms of informativeness between the Transmission-only and Transmission~+ Communication conditions only just (or not quite) meeting the 95\% criterion. This is likely to be related to a combination of two factors: a relatively small effect size combined with a relatively small sample size, with only ten chains---ten independent sampling units---in each condition. This is a general issue with iterated learning experiments that results from their resource-intensive nature, and our use of ten chains follows or exceeds norms in the field \parencite[e.g.,][]{Kempe:2015, Kirby:2008, Kirby:2015, Raviv:2018, Roberts:2018, Smith:2010, Tamariz:2015}. Nevertheless, our study suggests that future work in this area could benefit from increased power.}

\newmaterial{One important thing to note is that the participant population we draw from (native English speakers) is already accustomed to heterography and opacity; informative orthography might be even less forthcoming in other populations used to more transparent writing systems. This brings us to a much deeper issue with iterated learning experiments in general: We cannot avoid the fact that our participants come into the lab with prior linguistic baggage, whether that baggage is for the encoding of sound or the encoding of meaning. Ideally, our experiments would be performed with participants who have no writing experience at all, but since that would be very difficult to achieve, perhaps the second-best option is a participant population that is relatively open-minded to both types of writing systems. In this sense, our use of English speakers is actually quite appropriate, since the English writing system is neither fully phonographic nor fully logographic.}

@@

% \section{Acknowledgments}

% \noindent This work was funded by a Leverhulme Trust Research Project Grant (grant number: RPG-2020-034).

@@

>>>>

% temp for revision %%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{xcolor}
\newcommand\newmaterial[1]{\textcolor{black}{#1}}
\newcommand\secondrevision[1]{\textcolor{black}{#1}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\authorsnames{Jon W. Carr, Kathleen Rastle}

\authorsaffiliations{{Department of Psychology, Royal Holloway, University of London, England}}

\authornote{\textbf{Cite as:} Carr, J.\@ W., \& Rastle, K.\@ (2024). Why do languages tolerate heterography? An experimental investigation into the emergence of informative orthography. \textit{PsyArXiv}. Version~3. \url{https://doi.org/10.31234/osf.io/qnz5k}}

@@

Despite the benefits that heterography may provide in reading, it comes with two main costs. Firstly, by definition, heterography implies that a single sound can be spelled multiple ways. In English, the heterographic spellings <meat> and <meet> imply that /iː/ can be spelled <ea> or <ee>. Readers are therefore required to learn alternate spellings for a single sound, \newmaterial{resulting in longer learning periods and more difficult decoding \parencite{Reis:2020, Seymour:2003, Spencer:2003, Taylor:2011, Zhao:2018}}. Secondly, the arbitrary mapping between heterographic forms and meaning must also be learned. From the point of view of a modern English speaker, there is no intrinsic reason why \textit{meat} is spelled <ea> and \textit{meet} is spelled <ee>. Nevertheless, these arbitrary spelling distinctions must be learned if they are to be useful, and they presumably place an additional burden on reading and---perhaps even more-so---on writing \parencite{Frith:1979, Shankweiler:1992}.

@@

First, however, it is useful to consider the mechanisms by which selection could occur if it is indeed the case that heterography emerges for functional reasons. \textcite[pp.~325--326]{Berg:2021} outline two models of how a word might enter a state of heterography. The first model, \textit{the differentiation model} (Fig.~\ref{fig01}A), explains heterography through the creation of new orthographic forms. For example, the spelling <lite> for the word \textit{light} is frequently used in food products to mean light-in-calorific-weight; in British English, the spelling <cheque> (\newmaterial{perhaps influenced by French \textit{chèque}}) differentiates the bank draft from other meanings of the word \textit{check}; and the word \textit{byte} was a deliberate respelling of \textit{bite} to avoid accidental mutation into the closely related term \textit{bit} \parencite{Buchholz:1977}. Many monosyllabic words that are homophonous with common function words also tend to adopt alternate spellings, often by appending <e> or by doubling the final consonant: \textit{be}--\textit{bee}, \textit{but}--\textit{butt}, \textit{by}--\textit{bye}--\textit{buy}, \textit{for}--\textit{fore}--\textit{four}, \textit{in}--\textit{inn}, \textit{or}--\textit{oar}--\textit{ore}, \textit{so}--\textit{sew}, \textit{to}--\textit{too}--\textit{two}, \textit{we}--\textit{wee}. Differentiation is also common in surnames---\textit{Clarke}, \textit{Greene}, \textit{Wilde}; \textit{Carr}, \textit{Hogg}, \textit{Mann}---and trade names---\textit{Blu Tack}, \textit{Froot Loops}, \textit{Wite-Out} \parencite[§~6]{Carney:1994}. One important way in which differentiation can occur---especially in a language like English that has historically contained a lot of spelling variation \parencite{Nevalainen:2012, Stenroos:2016}---is by the conditioning of variant spellings on meaning; pairs like \textit{discreet}--\textit{discrete}, \textit{flour}--\textit{flower}, and \textit{plain}--\textit{plane}, which were once variant spellings of the same word, have taken on distinct meanings over time \parencite[§~5.4]{Carney:1994}. \textcite[p.~58]{Berg:2017} have referred to this as the ``functionalization of leftovers'': The spelling variants that survive are those that ``can find distributional or functional niches.''

@@

Our paper reports the results of two experiments---focusing on the differentiation and conservation models respectively---with the goal of demonstrating that functional heterography arises preferentially under a communicative need for disambiguation. All data and code is available from \url{https://osf.io/7auw6/}. \newmaterial{To increase the transparency of our work, we created a preregistration at \url{https://aspredicted.org/YDV_MXV}. Note, however, that due to the more exploratory nature of this project, our preregistration did not specify strong confirmatory hypotheses or precise statistical models, focusing instead on the research question, experimental conditions, general predictions, primary measurement constructs, sample size, and exclusion criteria.}

@@

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/fig07.pdf}}
	\vspace*{2pt}
	\caption{Results of Experiment~1. \textbf{A} Typological distribution by generation and condition over the four typological categories: holistic (H; purple), expressive (E; blue), redundant (R, green), and degenerate (D; yellow). \textbf{B} Proportion of systems classified as informative (holistic or expressive) by generation. The dots show the observed proportions and the curves show logistic regression models fit to the data. \textbf{C} Communicative cost by generation along with regression models fit to the data. \textbf{D} Transmission error by generation along with regression models fit to the data. The panels on the right show the posterior estimates of the slope ($\beta$) parameters by condition as well as the posterior differences between conditions. The green, blue, and red bars indicate respectively the 95\%, 90\%, and 85\% HDIs (credible intervals).}
	\label{fig07}
	\end{figure}

@@

\newmaterial{To analyze how informativeness changes over time and how this compares between the two conditions, we reduced the typological classifications into two broader categories: informative systems (i.e., holistic or expressive) and uninformative systems (i.e., redundant or degenerate). We then fit a Bayesian mixed-effects logistic regression model that predicts whether or not a system is informative as a function of generation with by-chain random slopes and intercepts,\footnote{Model: informative $\sim$ log(generation) + (1 + log(generation) | chain). Bernoulli likelihood with logit link function, weakly informative default priors, six chains of 12000 samples. All statistical models were fit using Bambi 0.13 \parencite{Capretto:2022}. Parameter estimates and diagnostics are provided in Appendix~B of the supplementary material.} which is the standard model structure used to analyze iterated learning experiments \parencite{Winter:2016}. Our Bayesian approach produces posterior estimates of two key parameters: $\alpha$, which represents the intercept of the regression model \secondrevision{(i.e., the model estimate of the dependent variable at Generation~0)}, and $\beta$, which represents its slope \secondrevision{(i.e., the model estimate of how much the dependent variable changes per generation)}. To determine if there is a statistical difference between conditions, we compute the difference in slopes, $\Delta(\beta) = \beta_\mathrm{comm} - \beta_\mathrm{trans}$, and check if this posterior difference satisfactorily rejects zero. \secondrevision{In other words, we test to see whether the dependent variable is changing over time more rapidly in one condition compared to the other.} Here we follow the convention that the 95\% highest density interval (HDI; the narrowest interval that contains 95\% of the posterior probability mass) should not include zero. We emphasize, however, that the posterior is a complete description of the evidence (given the data and model assumptions) and does not strictly need to be reduced to a binary yes/no decision. The results are shown in Fig.~\ref{fig07}B. In both conditions, there is a decrease in informativeness over time (the slopes are negative), but in the Transmission + Communication condition, the slope is shallower, suggesting that informativeness decreases more slowly in the presence of communicative pressure. The difference in slopes, $\Delta(\beta) = 4.53$ (95\%~HDI: 1.48, 7.88), clearly rejects zero, pointing to a meaningful difference between conditions.}

\newmaterial{One issue with the above approach is that collapsing the systems into binary categories (informative vs.\@ uninformative) results in a loss of information about \textit{how informative} the systems are. In addition, there was evidence to suggest that the model was a suboptimal description of the data, since there was also a difference in the $\alpha$ estimates (the intercepts\secondrevision{; Table~B1 in the supplementary material}), which should theoretically be the same (i.e., there should be no difference between conditions at Generation~0). We address these limitations with a second measure of informativeness, communicative cost \parencite{KempRegier:2012, Kemp:2018, Regier:2015}, which has previously been used in similar experimental studies \parencite{Carr:2020, Carstensen:2015, Smith:2020} and was proposed in our preregistration as the primary measure of informativeness.} Communicative cost is an information-theoretic measure that expresses how much information will be lost, on average, when a speaker/writer attempts to convey a meaning to a listener/reader using some shared signaling system. If the system contains no ambiguities (all meanings are expressed by unique signals), communicative cost will be zero~bits---that is, zero information will be lost during each attempt to communicate using that system. Communicative cost will take some larger value if the system contains ambiguity. It is given by $\sum_{m \in U} \mathrm{Pr}(m) ( -\log \mathrm{Pr}(m|s_m) )$, where $U$ is the universe of meanings that may be expressed, $\mathrm{Pr}(m)$ is the probability that a particular meaning would need to be expressed, and $\mathrm{Pr}(m|s_m)$ is the probability that a reader would infer meaning $m$ given that a writer produced signal $s$ for meaning $m$. In our case, $U$ is the set of nine alien objects, $\mathrm{Pr}(m)$ is set to $1/|U|$ (all objects need to be talked about with equal probability), and $\mathrm{Pr}(m|s_m)$ is given by $1/|M_s|$, where $M_s$ is the set of meanings labeled $s$ according to the system. \newmaterial{The results are shown in Fig.~\ref{fig07}C. We used the same mixed-effects linear regression model described above (except that the likelihood is now Gaussian). In line with the previous analysis, cost increases with generation in both conditions but increases more slowly under communicative pressure. However, the support for a difference between conditions was weaker under this more nuanced measure, $\Delta(\beta) = -0.17$ (95\%~HDI: $-0.36$, $0.02$); although we were not able to reject zero at the 95\% level, we were able to reject it at the 92\% level (92\%~HDI: $-0.35$, $-0.01$).}

\newmaterial{Aside from the informativeness of the orthographic systems, we also predicted in our preregistration that the systems would become easier to learn over time in both conditions, albeit for slightly different reasons. In Transmission-only, the orthographic system is expected to become increasingly learnable as it degenerates into a single, transparent suffix form. In Transmission~+ Communication, the system is expected to become more learnable as the unsystematic, holistic systems transform into other easier to learn systems (notably expressive systems, although---as noted already---expressive systems rarely emerged). Following prior work, we operationalized learnability as transmission error---the amount of error that the participant at Generation~$i$ made in reproducing the orthographic system that existed at Generation~$i-1$; transmission error is defined as the mean Levenshtein edit distance between the corresponding orthographic forms in consecutive generations \parencite[see e.g.,][]{Kirby:2008}. These results are plotted in Fig.~\ref{fig07}D. In both conditions, the estimates of the \textit{β} parameters are negative and clearly reject zero, suggesting that the systems do indeed become increasingly learnable over time \secondrevision{as hypothesized. Our analysis also suggested that there was little difference between the conditions in terms of how rapidly transmission error decreases over time (i.e., Δ(\textit{β}) is highly compatible with zero), although we did note that transmission error tended to be a little higher in the communicative condition.} This is to be expected because the goal in the Transmission-only condition is to reproduce the forms taught in training, whereas the goal in the communicative condition is to devise a system that permits accurate communication, which necessitates greater change to the system taught in training.}

@@

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/fig11.pdf}}
	\vspace*{2pt}
	\caption{Results of Experiment~2. \textbf{A} Typological distribution by generation and condition over the four typological categories: holistic (H; purple), expressive (E; blue), redundant (R, green), and degenerate (D; yellow). \textbf{B} Proportion of systems classified as informative (holistic or expressive) by generation. The dots show the observed proportions and the curves show logistic regression models fit to the data. \textbf{C} Communicative cost by generation along with regression models fit to the data. \textbf{D} Transmission error by generation along with regression models fit to the data. The panels on the right show the posterior estimates of the slope ($\beta$ and $\gamma$) parameters by condition as well as the posterior differences between conditions. $\beta$ represents the effect of epoch and $\gamma$ represents the effect of generation number within epoch. The green, blue, and red bars indicate respectively the 95\%, 90\%, and 85\% HDIs (credible intervals).}
	\label{fig11}
	\end{figure}

@@

Fig.~\ref{fig11}A plots the typological distributions by generation and condition. Initially, all systems are expressive, but the dominance of this category is gradually eroded over time, particularly during the second and third epochs once the spoken forms had become homophonous. Notably, however, the loss of expressive systems appears to be slower in Transmission~+ Communication, and redundant systems were also less popular under communicative pressure. \newmaterial{As in Experiment~1, we further collapsed the typological categories into two broader categories (informative vs.\@ uninformative) to analyze the trends over time. The results, shown in Fig.~\ref{fig11}B, show that the probability of a system being informative drops over time in both conditions (primarily as a function of epoch), but does so more slowly in the communicative condition. Although there was some weak evidence of a difference in epoch slopes ($\Delta(\beta) = 2.65$; 95\%~HDI: $-0.58$, $5.95$), we could not conclusively reject zero under this first measure of informativeness.}

\newmaterial{The results in terms of the preregistered measure of informativeness, communicative cost, are presented in Fig.~\ref{fig11}C. Here we find a nonzero effect of both epoch \secondrevision{($\beta$)} and generation \secondrevision{($\gamma$)} on cost, as well as a difference between conditions in terms of epoch: $\Delta(\beta) = -0.2$ (95\%~HDI: $-0.4$, $-0.003$). Like the previous measure, the $\gamma$ slopes were in close alignment\secondrevision{, so $\Delta(\gamma)$ is highly compatible with zero difference}. This suggests that the effect of generational turnover is very similar between conditions and that the difference between conditions is mostly driven by the increases in homophony induced in each epoch. The overall result is that, in Transmission~+ Communication, the increase in cost is linear across the nine generations, whereas in the Transmission-only condition, the increase in cost follows something more akin to a step function, with sudden increases in cost in response to each additional bout of homophony. In other words, in the Transmission-only condition, the orthographic systems respond rapidly to the changing spoken forms, while in the Transmission~+ Communication condition, the orthographic systems are more resistant to the homophony.}

@@

Although the emergent orthographies tended to be slightly more informative under communicative pressure, systematic differentiation was rare, unstable, and fleeting, be it through implicit generalization of the supplied variants or explicit innovation of new variants. This is \textit{not} because participants were unable to learn variant spellings; in many cases variant spellings were retained but ineffectually conditioned on shape. Nor was it because participants were unable to learn a system of color marking that is not expressed in the spoken language; we know from Experiment~2 that participants can learn and reproduce such systems. Instead, it seems that, in Experiment~1, differentiation could not get off the ground. We see two reasons for this. First, participants seemed disinclined to directly encode meaning in how they chose to spell, preferring instead to ``write by ear'' \parencite{Frith:1979}. When asked to type in a word for a pink pentagon called /buvɪkəʊ/, participants were inclined to type a sequence of graphemes that reflected the sound they heard, without encoding meaning. \newmaterial{This behavior might be connected to the concept of ``functional fixedness'' \parencite{German:2000}, which states that learners find it difficult to adduce a new function (e.g., writing meaning) when they are accustomed to another function (e.g., writing sound), which highlights a potentially important role for generational turnover in the development of writing systems, since new learners will be more receptive to new functions.} Second, even when participants \textit{did} appreciate the need to differentiate the written forms to be successful, they often appeared reluctant or unable to do so, perhaps because they viewed the spelling as immutable or because the problem of aligning with a partner---even in a~synchronous setting---was too difficult to overcome \newmaterial{without the ability to coordinate over an extended period of time}. It is notable, for example, that in the communication games, many attempts to differentiate using English color letters were not reciprocated.

@@

\newmaterial{During the review process, a concern was raised that our participants might not have fully understood the communicative nature of the task, thus explaining why we did not observe the emergence of systematic differentiation in spelling. Our position, as outlined above, was that the lack of differentiation was due to the very strong homophony pressure. To test whether the lack of informativeness might be attributed to the homophony and to check that the design of our experiment and implementation of the communicative pressure was sufficient to promote more informative systems, we ran an additional experiment. This experiment was identical to the communicative condition of Experiment~1, except that we removed the spoken forms (thereby replicating previous iterated learning experiments, which are generally only orthographic in nature; e.g., Beckner et al., 2017; Kirby et al., 2008, 2015) and altered the orthographic forms so that no homophony was implied in the spellings. This experiment is described in Appendix~C of the supplementary material, but in short, we observed much greater levels of innovation and informativeness \secondrevision{in this new experiment, with a clear statistical difference between it and the results of Experiment~1. Broadly, the effect of removing the homophony pressure was that the systems remained more informative compared to Experiment~1 (in terms of both the proportion of systems classified as informative and communicative cost).} There was also a commensurate increase in communicative success as a result of the emergence of these more informative systems (a point we return to shortly). Importantly, this suggests that the participants in Experiment~1 did indeed understand the communicative imperative, but nevertheless preferred their spellings to encode sound rather than meaning. Or, rather, the cultural evolutionary process ultimately tended to favor simplicity over informativeness in this particular domain.}

@@

\secondrevision{We emphasize, however, that we did not make a-priori predictions about which of the two mechanisms might represent a better theory of the emergence of informative heterography and our experiments were not designed to test the two theories against each other. Instead, we draw this conclusion on the basis that it was difficult for informative heterography to get off the ground in Experiment~1 due to the homophony present in the spoken form (as clarified by Experiment 3), while some degree of informative heterography did persist in Experiment 2 in the face of increasing homophony. A useful way that future work could explore this hypothesis would be to fit models of differentiation and conservation to historical data on spelling change to see which model offers a better fit to the data.}

@@

\newmaterial{The results of both experiments were relatively weak statistically, with the differences in terms of informativeness between the Transmission-only and Transmission~+ Communication conditions only just (or not quite) meeting the 95\% criterion. This is likely to be related to a combination of two factors: a relatively small effect size combined with a relatively small sample size, with only ten chains---ten independent sampling units---in each condition. \secondrevision{Our decision to run ten chains per condition was primarily based on norms in the field \parencite[e.g.,][]{Kempe:2015, Kirby:2008, Kirby:2015, Raviv:2018, Roberts:2018, Smith:2010, Tamariz:2015}, since we had little idea of what effect size we could expect to find when designing the studies. Nevertheless, the small effect sizes we observed do suggest some risk of type one error and future work with this paradigm would benefit from increasing the number of chains in light of these relatively small effect sizes}.}

\newmaterial{Lastly, one important thing to note is that the participant population we draw from (native English speakers) is already accustomed to heterography and opacity; informative orthography might be even less forthcoming in other populations used to more transparent writing systems. This brings us to a much deeper issue with iterated learning experiments in general: We cannot avoid the fact that our participants come into the lab with prior linguistic baggage, whether that baggage is for the encoding of sound or the encoding of meaning. Ideally, our experiments would be performed with participants who have no writing experience at all, but since that would be very difficult to achieve, perhaps the second-best option is a participant population that is relatively open-minded to both types of writing systems. In this sense, our use of English speakers is actually quite appropriate, since the English writing system is neither fully phonographic nor fully logographic.}

@@

\section{Acknowledgments}

\noindent This work was funded by a Leverhulme Trust Research Project Grant (grant number: RPG-2020-034) awarded to KR.

@@

% \pagestyle{empty}
